# Probability spaces {#introduction}

This chapter deals with measures and probability spaces. At the end of
the chapter, we look more closely at discrete probability spaces.

The students are expected to acquire the following knowledge:

**Theoretical**

- Use properties of probability to calculate probabilities.
- Combinatorics.
- Understanding of continuity of probability.


**R**

- Vectors and vector operations.
- For loop.
- Estimating probability with simulation.
- *sample* function.
- Matrices and matrix operations.


```{r, echo = FALSE}
togs   <- T
tog_ex <- T
```



<style>
.fold-btn { 
  float: right; 
  margin: 5px 5px 0 0;
}
.fold { 
  border: 1px solid black;
  min-height: 40px;
}
</style>

<script type="text/javascript">
$(document).ready(function() {
  $folds = $(".fold");
  $folds.wrapInner("<div class=\"fold-blck\">"); // wrap a div container around content
  $folds.prepend("<button class=\"fold-btn\">Unfold Solution</button>");  // add a button
  $(".fold-blck").toggle();  // fold all blocks
  $(".fold-btn").on("click", function() {  // add onClick event
    $(this).text($(this).text() === "Fold Solution" ? "Unfold Solution" : "Fold Solution");  // if the text equals "Fold", change it to "Unfold"or else to "Fold" 
    $(this).next(".fold-blck").toggle("linear");  // "swing" is the default easing function. This can be further customized in its speed or the overall animation itself.
  })
});
</script>


## Measure and probability spaces
```{exercise, name = "Completing a set to a sigma algebra"}
Let $\Omega = \{1,2,...,10\}$ and let $A = \{\emptyset, \{1\}, \{2\}, \Omega \}$.

a. Show that $A$ is not a sigma algebra of $\Omega$.
b. Find the minimum number of elements to complete A to a sigma algebra of
$\Omega$.

```

<div class="fold">
```{solution, echo = tog_ex}


a. $1^c = \{2,3,...,10\} \notin A \implies$ $A$ is not sigma algebra.
b. First we need the complements of all elements, so we need to add sets
$\{2,3,...,10\}$ and $\{1,3,4,...,10\}$. Next we need unions of all sets --
we add the set $\{1,2\}$. Again we need the complement of this set, so we add
$\{3,4,...,10\}$. So the minimum number of elements we need to add is 4.

```
</div>

```{exercise, name = "Diversity of sigma algebras"}
Let $\Omega$ be a set.

a. Find the smallest sigma algebra of $\Omega$.
b. Find the largest sigma algebra of $\Omega$.

```

<div class="fold">
```{solution, echo = togs}


a. $A = \{\emptyset, \Omega\}$
  
b. $2^{\Omega}$

```
</div>

```{exercise}
Find all sigma algebras for $\Omega = \{0, 1, 2\}$.

```

<div class="fold">
```{solution, echo = togs}


- $A = \{\emptyset, \Omega\}$
- $A = 2^{\Omega}$
- $A = \{\emptyset, \{0\}, \{1,2\}, \Omega\}$
- $A = \{\emptyset, \{1\}, \{0,2\}, \Omega\}$
- $A = \{\emptyset, \{2\}, \{0,1\}, \Omega\}$

```
</div>

```{exercise, name = "Difference between algebra and sigma algebra"}

Let $\Omega = \mathbb{N}$ and $\mathcal{A} = \{A \subseteq \mathbb{N}: 
A \text{ is finite or } A^c \text{ is finite.} \}$. Show that 
$\mathcal{A}$ is an algebra but not a sigma algebra.


```

<div class="fold">
```{solution, echo = togs}


- $\emptyset$ is finite so $\emptyset \in \mathcal{A}$.

- Let $A \in \mathcal{A}$ and $B \in \mathcal{A}$. If both are finite, then
their union is also finite and therefore in $\mathcal{A}$. Let at least 
one of them not
be finite. Then their union is not finite. But $(A \cup B)^c = A^c \cap B^c$.
And since at least one is infinite, then its complement is finite and the
intersection is too. So finite unions are in $\mathcal{A}$.

- Let us look at numbers $2n$. For any $n$, $2n \in \mathcal{A}$ 
as it is finite. But $\bigcup_{k = 1}^{\infty} 2n \notin \mathcal{A}$.


```
</div>

```{exercise, label = saequivalence}
We define $\sigma(X) = \cap_{\lambda \in I} S_\lambda$ to be a sigma algebra, generated by the set $X$, where $S_\lambda$ are all sigma algebras such that $X \subseteq S_\lambda$.
$S_\lambda$ are indexed by $\lambda \in I$.
Let $A, B \subseteq 2^{\Omega}$.
Prove that $\sigma(A) = \sigma(B) \iff A \subseteq \sigma(B) \land B \subseteq \sigma(A)$.
```

<div class="fold">
```{solution, echo = togs}
To prove the equivalence, we need to prove that the left hand side implies the right hand side and vice versa.

Proving $\sigma(A) = \sigma(B) \Rightarrow A \subseteq \sigma(B) \land B \subseteq \sigma(A)$:
we know $A \subseteq \sigma(A)$ is always true, so by substituting in $\sigma(B)$ from the left hand side equality we obtain $A \subseteq \sigma(B)$.
We obtain $B \subseteq \sigma(A)$ by symmetry.
This proves the implication.

Proving $A \subseteq \sigma(B) \land B \subseteq \sigma(A) \Rightarrow \sigma(A) = \sigma(B)$:
by definition of a sigma algebra, generated by a set, we have $\sigma(B) = \cap_{\lambda \in I} S_\lambda$ where $S_\lambda$ are all sigma algebras where $B \subseteq S_\lambda$.
But $\sigma(A)$ is one of $S_\lambda$, so we can write $\sigma(B) = \sigma(A) \cap \left(\cap_{\lambda \in I} S_\lambda \right)$, which implies $\sigma(B) \subseteq \sigma(A)$.
By symmetry, we have $\sigma(A) \subseteq \sigma(B)$.
Since $\sigma(A) \subseteq \sigma(B)$ and $\sigma(B) \subseteq \sigma(A)$, we obtain $\sigma(A) = \sigma(B)$, which proves the implication and completes the equivalence proof.
```
</div>


```{exercise, name = "Intro to measure"}

Take the measurable space $\Omega = \{1,2\}$, $F = 2^{\Omega}$. Which of the
following is a measure? Which is a probability measure?

a. $\mu(\emptyset) = 0$, $\mu(\{1\}) = 5$, $\mu(\{2\}) = 6$, $\mu(\{1,2\}) = 11$
b. $\mu(\emptyset) = 0$, $\mu(\{1\}) = 0$, $\mu(\{2\}) = 0$, $\mu(\{1,2\}) = 1$
c. $\mu(\emptyset) = 0$, $\mu(\{1\}) = 0$, $\mu(\{2\}) = 0$, $\mu(\{1,2\}) = 0$
d. $\mu(\emptyset) = 0$, $\mu(\{1\}) = 0$, $\mu(\{2\}) = 1$, $\mu(\{1,2\}) = 1$
e. $\mu(\emptyset)=0$, $\mu(\{1\})=0$, $\mu(\{2\})=\infty$, $\mu(\{1,2\})=\infty$

```

<div class="fold">
```{solution, echo = tog_ex}


  a. Measure. Not probability measure since $\mu(\Omega) > 1$.
  b. Neither due to countable additivity.
  c. Measure. Not probability measure since $\mu(\Omega) = 0$.
  d. Probability measure.
  e. Measure. Not probability measure since $\mu(\Omega) > 1$.

```
</div>

```{exercise}
Define a probability space that could be used to model the outcome of throwing two fair 6-sided dice.

```

<div class="fold">
```{solution, echo = togs}


- $\Omega = \{\{i,j\}, i = 1,...,6, j = 1,...,6\}$
  
- $F = 2^{\Omega}$
  
- $\forall \omega \in \Omega$, $P(\omega) = \frac{1}{6} \times \frac{1}{6} = \frac{1}{36}$

```
</div>

## Properties of probability measures

```{exercise}

A standard deck (52 cards) is distributed to two persons: 26 cards to each
person. All partitions are equally likely. Find the probability that:

a. The first person gets 4 Queens.
b. The first person gets at least 2 Queens.
c. <span style="color:blue"> R: Use simulation (_sample_) to check the above
answers. </span>

```

<div class="fold">
```{solution, echo = togs}


a. $\frac{\binom{48}{22}}{\binom{52}{26}}$
  
b. 1 - $\frac{\binom{48}{26} + 4 \times \binom{48}{25}}{\binom{52}{26}}$

c. For the simulation, let us represent cards with numbers from 1 to 52, and
let 1 through 4 represent Queens.

```
```{r, echo = togs, eval = togs}
set.seed(1)
cards <- 1:52
n     <- 10000
q4    <- vector(mode = "logical", length = n)
q2    <- vector(mode = "logical", length = n)

tmp <- vector(mode = "logical", length = n)
for (i in 1:n) {
  p1      <- sample(1:52, 26)
  q4[i]   <- sum(1:4 %in% p1) == 4
  q2[i]   <- sum(1:4 %in% p1) >= 2
}
sum(q4) / n
sum(q2) / n

```
</div>

```{exercise}
Let $A$ and $B$ be events with probabilities $P(A) = \frac{2}{3}$ and
$P(B) = \frac{1}{2}$.

a. Show that $\frac{1}{6} \leq P(A\cap B) \leq \frac{1}{2}$,
and give examples to show that both extremes are possible.
b. Find corresponding bounds for $P(A\cup B)$.
c. <span style="color:blue"> R: Draw samples from the examples and show the probability bounds of 
$P(A \cap B)$ </span>.

```

<div class="fold">
```{solution, echo = togs}

- From the properties of probability we have

\begin{equation}
  P(A \cup B) = P(A) + P(B) - P(A \cap B) \leq 1.
\end{equation}
  
From this follows

\begin{align}
  P(A \cap B) &\geq P(A) + P(B) -  1 \\
              &= \frac{2}{3} + \frac{1}{2} - 1 \\
              &= \frac{1}{6},
\end{align}

which is the lower bound for the intersection. Conversely, we have

\begin{equation}
  P(A \cup B) = P(A) + P(B) - P(A \cap B) \geq P(A).
\end{equation}

From this follows

\begin{align}
  P(A \cap B) &\leq P(B) \\
              &= \frac{1}{2},
\end{align}
  
which is the upper bound for the intersection.

For an example take a fair die. To achieve the lower bound let $A = \{3,4,5,6\}$
and $B = \{1,2,3\}$, then their intersection is $A \cap B = \{3\}$. To
achieve the upper bound take $A = \{1,2,3,4\}$ and $B = \{1,2,3\} $.


- For the bounds of the union we will use the results from the first part. 
Again from the properties of probability we have

\begin{align}
  P(A \cup B) &= P(A) + P(B) - P(A \cap B) \\
              &\geq P(A) + P(B) - \frac{1}{2} \\
              &= \frac{2}{3}.
\end{align}
  
Conversely

\begin{align}
  P(A \cup B) &= P(A) + P(B) - P(A \cap B) \\
              &\leq P(A) + P(B) - \frac{1}{6} \\
              &= 1.
\end{align}
  
Therefore $\frac{2}{3} \leq P(A \cup B) \leq 1$.


- We use _sample_ in R:

```
```{r, echo = togs, eval = togs}
set.seed(1)
n      <- 10000
samps  <- sample(1:6, n, replace = TRUE)

# lower bound
lb     <- vector(mode = "logical", length = n)
A      <- c(1,2,3)
B      <- c(3,4,5,6)
for (i in 1:n) {
  lb[i] <- samps[i] %in% A & samps[i] %in% B
}
sum(lb) / n

# upper bound
ub     <- vector(mode = "logical", length = n)
A      <- c(1,2,3)
B      <- c(1,2,3,4)
for (i in 1:n) {
  ub[i] <- samps[i] %in% A & samps[i] %in% B
}
sum(ub) / n
```
</div>

```{exercise}
A fair coin is tossed repeatedly. 

a. Show that, with probability one, a head turns up sooner or later. 

b. Show similarly that any given finite sequence of heads and
tails occurs eventually with probability one.
```

<div class="fold">
```{solution, echo = tog_ex}
\begin{align}
P(\text{no heads}) &= \lim_{n \rightarrow \infty} P(\text{no heads in first }n 
                                 \text{ tosses}) \\
                   &= \lim_{n \rightarrow \infty} \frac{1}{2^n} \\
                   &= 0.
\end{align}

For the second part, let us fix the given sequence of heads and tails of length 
$k$ as $s$. A probability that this happens in $k$ tosses is $\frac{1}{2^k}$.

\begin{align}
P(s \text{ occurs}) &= \lim_{n \rightarrow \infty} P(s \text{ occurs in first } 
                                                     nk \text{ tosses})
\end{align}

The right part of the upper equation is greater than if $s$ occurs either in the
first $k$ tosses, second $k$ tosses,..., $n$-th $k$ tosses. Therefore

\begin{align}
P(s \text{ occurs}) &\geq \lim_{n \rightarrow \infty} 
                  P(s \text{ occurs in first } n \text{ disjoint sequences of length } k) \\
                  &= \lim_{n \rightarrow \infty} 
                  (1 - P(s \text{ does not occur in first } n \text{ disjoint sequences})) \\
                  &= 1 -  \lim_{n \rightarrow \infty} P(s \text{ does not occur in first } n \text{ disjoint sequences}) \\
                  &= 1 - \lim_{n \rightarrow \infty} (1 - \frac{1}{2^k})^n \\ 
                  &= 1.
\end{align}

```
</div>

```{exercise}
An Erdos-Renyi random graph $G(n,p)$ is a model with $n$ nodes, where
each pair of nodes is connected with probability $p$.


a. Calculate the probability that there exists a node that is not
connected to any other node in $G(4,0.6)$.
b. Show that the upper bound for the probability that there exist 2 nodes that are not connected to any other node for an arbitrary $G(n,p)$ is $\binom{n}{2} (1-p)^{2n - 3}$.
c. <span style="color:blue">R: Estimate the probability from the first point using simulation.</span>
  

```

<div class="fold">
```{solution, echo = togs}


a. Let $A_i$ be the event that the $i$-th node is not connected to
any other node. Then our goal is to calculate $P(\cup_{i=1}^n A_i)$.
Using the inclusion-exclusion principle, we get

\begin{align}
  P(\cup_{i=1}^n A_i) &= \sum_i A_i - \sum_{i<j} P(A_i \cap A_j) +
    \sum_{i<j<k} P(A_i \cap A_j \cap A_k) - P(A_1 \cap A_2 \cap A_3 \cap A_4) \\
                      &=4 (1 - p)^3 - \binom{4}{2} (1 - p)^5 +
                        \binom{4}{3} (1 - p)^6 - (1 - p)^6 \\
                      &\approx 0.21.
\end{align}
  
b. Let $A_{ij}$ be the event that nodes $i$ and $j$ are not connected to any other node. We are interested in $P(\cup_{i<j}A_{ij})$. By using Boole`s inequality, we get

\begin{align}
P(\cup_{i<j}A_{ij}) \leq \sum_{i<j} P(A_{ij}).
\end{align}

What is the probability of $A_{ij}$? There need to be no connections to the $i$-th node to the remaining nodes (excluding $j$), the same for the $j$-th node, and there can be no connection between them. Therefore

\begin{align}
P(\cup_{i<j}A_{ij}) &\leq \sum_{i<j} (1 - p)^{2(n-2) + 1} \\
                    &= \binom{n}{2} (1 - p)^{2n - 3}.
\end{align}


```
```{r, echo = togs, eval = togs}
set.seed(1)
n_samp <- 100000
n      <- 4
p      <- 0.6
conn_samp <- vector(mode = "logical", length = n_samp)
for (i in 1:n_samp) {
  tmp_mat                     <- matrix(data = 0, nrow = n, ncol = n)
  samp_conn                   <- sample(c(0,1), 
                                        choose(4,2), 
                                        replace = TRUE, 
                                        prob    = c(1 - p, p))
  tmp_mat[lower.tri(tmp_mat)] <- samp_conn
  tmp_mat[upper.tri(tmp_mat)] <- t(tmp_mat)[upper.tri(t(tmp_mat))]
  not_conn                    <- apply(tmp_mat, 1, sum)
  if (any(not_conn == 0)) {
    conn_samp[i] <- TRUE
  } else {
    conn_samp[i] <- FALSE
  }
}
sum(conn_samp) / n_samp
```
</div>

## Discrete probability spaces
```{exercise, label = binomialpmf}
Show that the standard measurable space on $\Omega = \{0,1,...,n\}$ equipped with binomial measure is a discrete probability space.

a. Define another probability measure on this measurable space.
b. Show that for $n=1$ the binomial measure is the same as the Bernoulli measure.
c. <span style="color:blue">R: Draw 1000 samples from the binomial distribution $p=0.5$, $n=20$ (_rbinom_) and compare relative frequencies with theoretical probability measure</span>.

```

<div class="fold">
```{solution, echo = tog_ex}
We need to show that the terms of 
$\sum_{k=0}^n \binom{n}{k} p^k (1 - p)^{n - k}$ sum to 1. For that we use the
binomial theorem $\sum_{k=0}^n \binom{n}{k} x^k y^{n-k} = (x + y)^n$. So

\begin{equation}
  \sum_{k=0}^n \binom{n}{k} p^k (1 - p)^{n - k} = (p + 1 - p)^n = 1.
\end{equation}

- $P(\{k\}) = \frac{1}{n + 1}$.
- When $n=1$ then $k \in \{0,1\}$. Inserting $n=1$ into the binomial measure, 
we get $\binom{1}{k}p^k (1-p)^{1 - k}$. Now 
$\binom{1}{1} = \binom{1}{0} = 1$, so the measure is $p^k (1-p)^{1 - k}$, 
which is the Bernoulli measure.

```
```{r, echo = tog_ex, message = FALSE, warning=FALSE, eval = tog_ex}
set.seed(1)
library(ggplot2)
library(dplyr)
bin_samp <- rbinom(n = 1000, size = 20, prob = 0.5)
bin_samp <- data.frame(x = bin_samp) %>%
  count(x) %>%
  mutate(n = n / 1000, type = "empirical_frequencies") %>%
  bind_rows(data.frame(x = 0:20, n = dbinom(0:20, size = 20, prob = 0.5), type = "theoretical_measure"))

bin_plot <- ggplot(data = bin_samp, aes(x = x, y = n, fill = type)) +
  geom_bar(stat="identity", position = "dodge")
plot(bin_plot)
```
</div>


```{exercise, label = geopoispmf}
Show that the standard measurable space on $\Omega = \{0,1,...,\infty\}$
  
   
a. equipped with geometric measure is a discrete probability space,
b. equipped with Poisson measure is a discrete probability space.
c. Define another probability measure on this measurable space.
d. <span style="color:blue">R: Draw 1000 samples from the Poisson distribution $\lambda = 10$ (_rpois_) and compare relative frequencies with theoretical probability measure</span>.

```

<div class="fold">
```{solution, echo = togs}


a. $\sum_{k = 0}^{\infty} p(1 - p)^k = p \sum_{k = 0}^{\infty} (1 - p)^k =
  p \frac{1}{1 - 1 + p} = 1$. We used the formula for geometric series.

b. $\sum_{k = 0}^{\infty} \frac{\lambda^k e^{-\lambda}}{k!} = 
  e^{-\lambda} \sum_{k = 0}^{\infty} \frac{\lambda^k}{k!} =
  e^{-\lambda} e^{\lambda} = 1.$ We used the Taylor expansion of
  the exponential function.
  
c. Since we only have to define a probability
measure, we could only assign probabilities that sum to one to a finite 
number of events in $\Omega$, and probability zero to the other infinite 
number of events. However to make this solution more educational, we will 
try to find a measure that assigns a non-zero probability to all events in 
$\Omega$. A good start for this would be to find a converging infinite series,
as the probabilities will have to sum to one. One simple converging series is 
the geometric series $\sum_{k=0}^{\infty} p^k$ for $|p| < 1$. Let us choose an 
arbitrary $p = 0.5$. Then $\sum_{k=0}^{\infty} p^k = \frac{1}{1 - 0.5} = 2$.
To complete the measure, we have to normalize it, so it sums to one,
therefore $P(\{k\}) = \frac{0.5^k}{2}$ is a probability measure on $\Omega$.

We could make it even more difficult by making this measure dependent
on some parameter $\alpha$, but this is out of the scope of this introductory
chapter.

  
```
```{r, echo = togs, message = FALSE, warning=FALSE, eval = togs}
set.seed(1)
pois_samp <- rpois(n = 1000, lambda = 10)
pois_samp <- data.frame(x = pois_samp) %>%
  count(x) %>%
  mutate(n = n / 1000, type = "empirical_frequencies") %>%
  bind_rows(data.frame(x = 0:25, n = dpois(0:25, lambda = 10), type = "theoretical_measure"))

pois_plot <- ggplot(data = pois_samp, aes(x = x, y = n, fill = type)) +
  geom_bar(stat="identity", position = "dodge")
plot(pois_plot)
```
</div>

```{exercise}


a. Define a probability measure on $(\Omega = \mathbb{Z}, 2^{\mathbb{Z}})$.

b. Define a probability measure such that $P(\omega) > 0, \forall \omega \in \Omega$.

c. <span style="color:blue">R: Implement a random generator that will generate 
samples with the relative frequency that corresponds to your probability
measure. Compare relative frequencies with theoretical probability 
measure </span>.

```

<div class="fold">
```{solution, echo = togs}


a. $P(0) = 1, P(\omega) = 0, \forall \omega \neq 0$.
b. $P(\{k\}) = \sum_{k = -\infty}^{\infty} \frac{p(1 - p)^{|k|}}{2^{1 - 1_0(k)}}$,
  where $1_0(k)$ is the indicator function, which equals to one if $k$ is 0, and
  equals to zero in every other case.

  
```
```{r, echo = togs, message = FALSE, warning=FALSE, eval = togs}
n          <- 1000
geom_samps <- rgeom(n, prob = 0.5)
sign_samps <- sample(c(FALSE, TRUE), size = n, replace = TRUE)
geom_samps[sign_samps] <- -geom_samps[sign_samps]
my_pmf <- function (k, p) {
  indic         <- rep(1, length(k))
  indic[k == 0] <- 0
  return ((p * (1 - p)^(abs(k))) / 2^indic)
}
geom_samps <- data.frame(x = geom_samps) %>%
  count(x) %>%
  mutate(n = n / 1000, type = "empirical_frequencies") %>%
  bind_rows(data.frame(x = -10:10, n = my_pmf(-10:10, 0.5), type = "theoretical_measure"))

geom_plot <- ggplot(data = geom_samps, aes(x = x, y = n, fill = type)) +
  geom_bar(stat="identity", position = "dodge")
plot(geom_plot)


```
</div>

```{exercise}
Define a probability measure on $\Omega = \{1,2,3,4,5,6\}$ with parameter $m 
\in \{1,2,3,4,5,6\}$, so that the probability of outcome at distance $1$ from
$m$ is half of the probability at distance $0$, at distance $2$ is half of
the probability at distance $1$, etc. <span style="color:blue">R: 
Implement a random generator that will generate 
samples with the relative frequency that corresponds to your probability
measure. Compare relative frequencies with theoretical probability 
measure </span>. 

```

<div class="fold">
```{solution, echo = togs}


- $P(\{k\}) = \frac{\frac{1}{2}^{|m - k|}}{\sum_{i=1}^6 \frac{1}{2}^{|m - i|}}$

  
```
```{r, echo = togs, message = FALSE, warning=FALSE, eval = togs}
n          <- 10000
m <- 4
my_pmf <- function (k, m) {
  denom <- sum(0.5^abs(m - 1:6))
  return (0.5^abs(m - k) / denom)
}
samps <- c()
for (i in 1:n) {
  a <- sample(1:6, 1)
  a_val <- my_pmf(a, m)
  prob  <- runif(1)
  if (prob < a_val) {
    samps <- c(samps, a)
  }
}
samps <- data.frame(x = samps) %>%
  count(x) %>%
  mutate(n = n / length(samps), type = "empirical_frequencies") %>%
  bind_rows(data.frame(x = 1:6, n = my_pmf(1:6, m), type = "theoretical_measure"))

my_plot <- ggplot(data = samps, aes(x = x, y = n, fill = type)) +
  geom_bar(stat="identity", position = "dodge")
plot(my_plot)
```
</div>

