# Convergence of random variables {#crv}

This chapter deals with convergence of random variables.

The students are expected to acquire the following knowledge:

**Theoretical**

- Finding convergences of random variables.



<style>
.fold-btn { 
  float: right; 
  margin: 5px 5px 0 0;
}
.fold { 
  border: 1px solid black;
  min-height: 40px;
}
</style>

<script type="text/javascript">
$(document).ready(function() {
  $folds = $(".fold");
  $folds.wrapInner("<div class=\"fold-blck\">"); // wrap a div container around content
  $folds.prepend("<button class=\"fold-btn\">Unfold</button>");  // add a button
  $(".fold-blck").toggle();  // fold all blocks
  $(".fold-btn").on("click", function() {  // add onClick event
    $(this).text($(this).text() === "Fold" ? "Unfold" : "Fold");  // if the text equals "Fold", change it to "Unfold"or else to "Fold" 
    $(this).next(".fold-blck").toggle("linear");  // "swing" is the default easing function. This can be further customized in its speed or the overall animation itself.
  })
});
</script>


```{r, echo = FALSE, warning = FALSE, message = FALSE}
togs <- TRUE
library(ggplot2)
library(dplyr)
library(reshape2)
library(tidyr)
# togs <- FALSE
```


```{exercise}
Let $X_1$, $X_2$,..., $X_n$ be a sequence of Bernoulli random variables. Let $Y_n = \frac{X_1 + X_2 + ... + X_n}{n^2}$. Show that this sequence converges point-wise to the zero random variable.
<span style="color:blue">R: Use a simulation to check your answer.</span>

```
<div class="fold">
```{solution, echo = togs}


Let $\epsilon$ be arbitrary. We need to find such $n_0$, that for every $n$ greater than $n_0$ $|Y_n| < \epsilon$ holds.

\begin{align}
  |Y_n| &= |\frac{X_1 + X_2 + ... + X_n}{n^2}|  \\
        &\leq |\frac{n}{n^2}| \\
        &= \frac{1}{n}.
\end{align}
So we need to find such $n_0$, that for every $n > n_0$ we will have $\frac{1}{n} < \epsilon$. So $n_0 > \frac{1}{\epsilon}$.


```
```{r, echo = togs, message = FALSE, warning=FALSE}
x <- 1:1000
X <- matrix(data = NA, nrow = length(x), ncol = 100)
y <- vector(mode = "numeric", length = length(x))
for (i in 1:length(x)) {
  X[i, ] <- rbinom(100, size = 1, prob = 0.5)
}
X <- apply(X, 2, cumsum)
tmp_mat <- matrix(data = (1:1000)^2, nrow = 1000, ncol = 100)
X <- X / tmp_mat
y <- apply(X, 1, mean)
ggplot(data.frame(x = x, y = y), aes(x = x, y = y)) +
  geom_line()
```
</div>

```{exercise}
Let $\Omega = [0,1]$ and let $X_n$ be a sequence of random variables, defined as
\begin{align}
  X_n(\omega) = \begin{cases}
    \omega^3, &\omega = \frac{i}{n}, &0 \leq i \leq 1 \\
    1, & \text{otherwise.}
  \end{cases}
\end{align}
Show that $X_n$ converges almost surely to $X \sim \text{Uniform}(0,1)$.
```
<div class="fold">
```{solution, echo = togs}


We need to show $P(\{\omega: X_n(\omega) \rightarrow X(\omega)\}) = 1$. 

Let $\omega \neq \frac{i}{n}$. Then for any $\omega$, $X_n$ converges pointwise to $X$:
\begin{align}
  X_n(\omega) = 1 \implies |X_n(\omega) - X(s)| = |1 - 1| < \epsilon.
\end{align}
The above is independent of $n$. Since there are countably infinite number of elements in the complement ($\frac{i}{n}$), the probability of this set is 1.
```
</div>


```{exercise}
Borrowed from Wasserman. Let $X_n \sim \text{N}(0, \frac{1}{n})$ and let $X$ be a random variable with CDF
\begin{align}
 F_X(x) = \begin{cases}
  0, &x < 0 \\
  1, &x \geq 0.
 \end{cases}
\end{align}
Does $X_n$ converge to $X$ in distribution? How about in probability? Prove or disprove these statement.
<span style="color:blue">R: Plot the CDF of $X_n$ for $n = 1, 2, 5, 10, 100, 1000$.</span>

```
<div class="fold">
```{solution, echo = togs}
Let us first check convergence in distribution.
\begin{align}
 \lim_{n \rightarrow \infty} F_{X_n}(x) &= \lim_{n \rightarrow \infty} \phi (\sqrt(n) x).
\end{align}
We have two cases, for $x < 0$ and $x > 0$. We do not need to check for $x = 0$, since $F_X$ is not continuous in that point.
\begin{align}
  \lim_{n \rightarrow \infty} \phi (\sqrt(n) x) = \begin{cases}
  0, & x < 0 \\
  1, & x > 0.
\end{cases}
\end{align}
This is the same as $F_X$.

Let us now check convergence in probability. Since $X$ is a point-mass distribution at zero, we have
\begin{align}
 \lim_{n \rightarrow \infty} P(|X_n| > \epsilon) &= \lim_{n \rightarrow \infty} (P(X_n > \epsilon) + P(X_n < -\epsilon)) \\
&= \lim_{n \rightarrow \infty} (1 - P(X_n < \epsilon) + P(X_n < -\epsilon)) \\
&= \lim_{n \rightarrow \infty} (1 - \phi(\sqrt{n} \epsilon) + \phi(- \sqrt{n} \epsilon)) \\
&= 0.
\end{align}


```
```{r, echo = togs, message = FALSE, warning=FALSE}
n <- c(1,2,5,10,100,1000)
ggplot(data = data.frame(x = seq(-5, 5, by = 0.01)), aes(x = x)) +
  stat_function(fun = pnorm, args = list(mean = 0, sd = 1/1), aes(color = "sd = 1/1")) +
  stat_function(fun = pnorm, args = list(mean = 0, sd = 1/2), aes(color = "sd = 1/2")) +
  stat_function(fun = pnorm, args = list(mean = 0, sd = 1/5), aes(color = "sd = 1/5")) +
  stat_function(fun = pnorm, args = list(mean = 0, sd = 1/10), aes(color = "sd = 1/10")) +
  stat_function(fun = pnorm, args = list(mean = 0, sd = 1/100), aes(color = "sd = 1/100")) +
  stat_function(fun = pnorm, args = list(mean = 0, sd = 1/1000), aes(color = "sd = 1/1000")) +
  stat_function(fun = pnorm, args = list(mean = 0, sd = 1/10000), aes(color = "sd = 1/10000"))
```
</div>


```{exercise}
Let $X_i$ be i.i.d. and $\mu = E(X_1)$. Let variance of $X_1$ be finite. Show that the mean of $X_i$, $\bar{X}_n = \frac{1}{n}\sum_{i=1}^n X_i$ converges in quadratic mean to $\mu$. 

```
<div class="fold">
```{solution, echo = togs, label = "vardecomp"}
\begin{align}
  \lim_{n \rightarrow \infty} E(|\bar{X_n} - \mu|^2) &= \lim_{n \rightarrow \infty} E(\bar{X_n}^2 - 2 \bar{X_n} \mu + \mu^2) \\
  &= \lim_{n \rightarrow \infty} (E(\bar{X_n}^2) - 2 \mu E(\frac{\sum_{i=1}^n X_i}{n}) + \mu^2) \\
  &= \lim_{n \rightarrow \infty} E(\bar{X_n})^2 + \lim_{n \rightarrow \infty} Var(\bar{X_n}) - 2 \mu^2 + \mu^2 \\
  &= \lim_{n \rightarrow \infty} \frac{n^2 \mu^2}{n^2} + \lim_{n \rightarrow \infty} \frac{\sigma^2}{n} - \mu^2 \\
  &= \mu^2 - \mu^2 + \lim_{n \rightarrow \infty} \frac{\sigma^2}{n} \\
  &= 0.
\end{align}



```
</div>
